---
title: "STA221"
author: "Neil Montgomery"
date: "Last edited: `r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output: 
  beamer_presentation:
    keep_tex: TRUE
    incremental: TRUE
#    df_print: tibble
    fig_caption: FALSE
classoption: aspectratio=169
header-includes:
- \newcommand{\ve}{\varepsilon}
- \newcommand{\dbar}[1]{\overline{\overline{#1}}}
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE,
                      dev='pdf', fig.width=5, fig.asp=0.618, fig.align='center')
options(tibble.width=70, scipen = 999, tibble.print_min=5, show.signif.stars = FALSE)
library(tidyverse)
```

# experiments with two factors

```{r}
burn <- read.delim("chelating.txt", check.names = FALSE, 
                   colClasses = c("factor", "factor", "numeric"))
```


## fire retardant example

There aren't many datasets that come with Chapter 26, so I found another to use as a motivating example.

In this dataset a new fire retardant treatment of cotton fabric is being tested. The name of this factor is `Bath`. There are two levels names `I` and `II`, which stand for "received the treatment" and "did not receive the new treatment" respectively. 

(It's called "Bath" because the fabric is bathed in the treatment solution.)

\pause The amount of time it takes each cloth sample to start to burn is recorded.

Here is a numerical summary of the results:

```{r, results='asis'}
library(xtable)
burn %>% 
  group_by(Bath) %>% 
  summarize(n=n(), Means=mean(`Burn-Times`), SD=sd(`Burn-Times`)) %>% 
  xtable(., digits=3) %>% print.xtable(., include.rownames=FALSE, comment=FALSE)
```

## but there is also another variable

The new treatment works. But there is also the matter of the efficiency with which the treatment can be applied.

There is another factor variable in this dataset: the number of "launderings", which is the way a retardant treatment is applied.

This variable has two levels named `5` and `10`, corresponding to the actual number of launderings.

Here is a summary of the results with respect to this variable:

```{r, results='asis'}
burn %>% 
  group_by(Launderings) %>% 
  summarize(n=n(), Means=mean(`Burn-Times`), SD=sd(`Burn-Times`)) %>% 
  xtable(., digits=3) %>% print.xtable(., include.rownames=FALSE, comment=FALSE)
```

## the effect of both variables

The new treatment is better than the old treatment (all else being equal). More laundering is better than less laundering (all else being equal).

But what about the combination of both variables? There are a few different ways they might combine to affect the results. We'll be concerned with these possibilities:

\pause Their effects could simply be \textit{additive}.

\pause The effect of one factor could depend on the level of the other factor. In this case we say there is an \textit{interaction} between the two factor variables.

\pause A good graphical method to evaluate the relationship is called an \textit{interaction plot}.

\pause To illustrate, I have simulated a dataset with two variables. `Factor1` has levels `A` and `B` while `Factor2` has levels `C` and `D`. 

## interaction plot example  1a - additive

```{r}
two_factor <- expand.grid(Factor1=rep(LETTERS[1:2], 8), 
                          Factor2=rep(LETTERS[3:4], 8)) %>% 
  mutate(y1 = as.numeric(Factor1)*1.5 + 
           as.numeric(Factor2)*0.5 + rnorm(nrow(.), 0, 0.5),
         y2 = as.numeric(Factor1)*1.5 + 
           as.numeric(Factor2)*0.5 + as.numeric(Factor1)*as.numeric(Factor2)*1.75+ rnorm(nrow(.), 0, 0.25),
         y3 = as.numeric(Factor1)*1.5 + 
           as.numeric(Factor2)*0.5 - as.numeric(Factor1)*as.numeric(Factor2) + rnorm(nrow(.), 0, 0.25))

two_factor %>% 
  ggplot(aes(x=Factor1, y=y1, colour=Factor2, group=Factor2)) + 
  stat_summary(fun.y=mean, geom="point") + 
  stat_summary(fun.y=mean, geom="line") + 
    ylab("means of the response variable") + 
  ggtitle("Effects are additive")
```

## interaction plot example  1b - additive

```{r}
two_factor %>% 
  ggplot(aes(x=Factor2, y=y1, colour=Factor1, group=Factor1)) + 
  stat_summary(fun.y=mean, geom="point") + 
  stat_summary(fun.y=mean, geom="line") + 
  ylab("means of the response variable") + 
  ggtitle("Doesn't matter which variable goes where")
```

## interaction plot example  2a - positive interaction

```{r}
two_factor %>% 
  ggplot(aes(x=Factor1, y=y2, colour=Factor2, group=Factor2)) + 
  stat_summary(fun.y=mean, geom="point") + 
  stat_summary(fun.y=mean, geom="line") + 
    ylab("means of the response variable") + 
  ggtitle("Positive interaction")
```

## interaction plot example  2b - positive interaction

```{r}
two_factor %>% 
  ggplot(aes(x=Factor2, y=y2, colour=Factor1, group=Factor1)) + 
  stat_summary(fun.y=mean, geom="point") + 
  stat_summary(fun.y=mean, geom="line") + 
    ylab("means of the response variable") + 
  ggtitle("Variables switched") 
```

## interaction plot example 3a - negative interaction

```{r}
two_factor %>% 
  ggplot(aes(x=Factor1, y=y3, colour=Factor2, group=Factor2)) + 
  stat_summary(fun.y=mean, geom="point") + 
  stat_summary(fun.y=mean, geom="line") + 
    ylab("means of the response variable") + 
  ggtitle("Negative interaction")
```


## interaction plot example 3a - negative interaction

```{r}
two_factor %>% 
  ggplot(aes(x=Factor2, y=y3, colour=Factor1, group=Factor1)) + 
  stat_summary(fun.y=mean, geom="point") + 
  stat_summary(fun.y=mean, geom="line") + 
    ylab("means of the response variable") + 
  ggtitle("Variables switched")
```

## interaction plot of the fire retardant data

```{r}
burn %>% 
  ggplot(aes(x=Launderings, y=`Burn-Times`, colour=Bath, group=Bath)) + 
  stat_summary(fun.y=mean, geom="point") + 
  stat_summary(fun.y=mean, geom="line") 
```

## fire data - variables switched

```{r}
burn %>% 
  ggplot(aes(x=Bath, y=`Burn-Times`, colour=Launderings, group=Launderings)) + 
  stat_summary(fun.y=mean, geom="point") + 
  stat_summary(fun.y=mean, geom="line") 
```

## tentative conclusion from fire data | new models to consider

It seems that the new treatment is just as good, even after only 5 launderings, so it is also more efficient.

\pause Here are the models we will consider:

$$y_{ijk} = \mu + \tau_i + \gamma_j + \ve_{ijk} \qquad \text{ no interaction assumed}$$

$$y_{ijk} = \mu + \tau_i + \gamma_j + (\tau\gamma)_{ij} +  \ve_{ijk} \qquad \text{ general model}$$

\pause $\mu$ is the grand, overall average.

\pause The $\tau_i$ are the effects of the levels of the first factor.

\pause The $\gamma_j$ are the effects of the levels of the second factor.

\pause The $(\tau\gamma)_{ij}$ are the effects of all combinations of the levels of hte two factors.

\pause $\ve_{ijk}$ is random noise, assumed to be $N(0,\sigma)$.

## the analysis

We have new sums of squares decompositions. The details can get brutal - so we'll stay symbolic. 

We'll call the factor variables `A` and `B`, with $I$ and $J$ levels respectively.

In the additive case we get, assuming a common sample size of $n$ for each combination of factor levels (strongly encouraged in practice!):

$$SS_{Total} =  SS_A + SS_B + SS_{Error}$$

\pause Everything has $\chi^2$ distributions. The degrees of freedom add up (N is the grand sample size):

$$N-1 = (I-1) + (J-1) + (N-I-J+1)$$

\pause ~~~When the experiment is \textit{balanced} (strongly encouraged!) with $n$ in each combination of levels, the error degrees of freedom simplifies to $IJ(n-1)$.~~~

## fire example - no interaction (?!)

```{r}
burn %>% 
  aov(`Burn-Times` ~ Bath + Launderings, data=.) %>% summary
```

